Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2020-08-20 2.background/2.RV_and_RS.tex
addded some exercises
to Chapter 2.

>>>>

\section{Exercises}

\fbox{ TO DO: Verify and add}

\begin{enumerate}

 \item (Inspired by \cite{resnick1973almost}) Let $\epsilon(i)\sim F,\ i=1,2,\cdots$ have arbitrary dependence 
 and $u_p = F^{\leftarrow}(1-1/p)$.

 {\bf (a)}   If for some $\delta_p\to 0$,
 we have $\sum_{p} \overline{F}((1+\delta_p)u_p) <\infty$, then show that
 $$
 \limsup_{p\to\infty} \frac{M_p}{u_p} \le 1,\ \ \mbox{ almost surely,} 
 $$
 where $M_p:= \max_{i=1,\cdots,p} \epsilon(i)$.\\

 {\bf (b)} Show that the condition of part {\bf (a)} holds for the Generalized AGG$(\nu),\ \nu>0$ distributions.\\

 {\em Hint:} In part {\bf (a)}, argue that the events $\{ M_p > (1+\delta_p) u_p,\ \mbox{ infinitely often}\}$ and
  $\{\epsilon(p) > )(1+\delta_p)u_p, \ \mbox{infinitely often}\}$ are equal. Appeal to the Borel Zero-One Law.\\

 \item  Show that independent realizations from a Generalized AGG$(\nu),\ \nu>0$ distribution
  are {\em almost surely stable}. That is, let $\epsilon(i)$'s be independent with distribution function $F$ as in Example \ref{exmp:AGG}.\\

{\bf (a)} Show that there exists a sequence $\delta_p\to 0$, such that 
$$
\sum_p \overline F((1+\delta_p)u_p) <\infty\ \ \mbox{ and }\ \  \sum_p F^p((1-\delta_p)u_p) <\infty.
$$
 {\bf (b)} By appealing to the Borel Zero-One law, conclude that in this case
 $$
 \frac{1}{u_p} M_p {\longrightarrow} 1,\ \ \mbox{ almost surely}.
 $$

{\em Comment:} 
\cite{resnick1973almost} provide a general necessary and sufficient condition for almost sure stability of independent maxima.\\

 \item Prove Gnedenko's result in part (2) of Proposition \ref{prop:rapid-varying-tails}.\\

\item Suppose that $F$ is heavy-tailed, i.e., $\overline F(x) \sim x^{-\alpha} L(x),$ as $x\to\infty$, for some $\alpha>0$ and a
slowly varying function $L$.  That is, $L(tx)/L(t)\to 1$, as $t\to\infty$, for all $x>0$.\\

{\bf (a)} Show that if $u_p$ is such that $p\overline F(u_p)\to 1$ as $p\to\infty$, then $p \overline F(x u_p)\to x^{-\alpha}$.\\

{\bf (b)} Let $\epsilon(i)$'s be independent realizations from $F$. With $u_n$ as in part {\bf (a)}, show that
$$ 
\frac{M_p}{u_p} \stackrel{d}{\longrightarrow} Z,\ \ \mbox{ as }p\to\infty,
$$ 
where $M_p = \max_{i=1,\cdots,p} \epsilon(i)$ and $\P(Z\le x) = e^{-1/x^\alpha},\ x>0$.\\

{\em Comment:} In the heavy-tailed case, rather than concentration of maxima (relative stability),
we encounter {\em dispersion of maxima} where under rescaling $M_p$ converges in distribution to a proper random variable.\\

\item Let  $M_p,\ p=1,2,\cdots$  be an infinite sequence of random variables such that
$$
\frac{M_p}{u_p} \stackrel{\P}{\longrightarrow} 1\ \ \mbox{ and }\ \ \frac{M_p}{v_p} \stackrel{d}{\longrightarrow} \xi,
$$
for some deterministic positive sequences  $u_p$ and $v_p$ and some non-zero random variable $\xi$.

Show that $\lim_{p\to\infty} u_p/v_p = c$ and $\xi = c$, almost surely.\\

{\em Hint:} Using Skorokhod's theorem there exist $\widetilde M_p$ and $\widetilde \xi$ on another probability space
such that $M_p/v_p\stackrel{d}{=}\widetilde M_p/v_p \stackrel{a.s.}{\to} \widetilde \xi \stackrel{d}{=}\xi$.
Then, recall the following criterion.

\begin{lemma} We have $\eta_p\stackrel{\P}{\to} \eta,\ p\to\infty$ if and only if, for every sequence $p_n\to \infty$, there is a further
subsequence $\{\widetilde p_n\}\subset\{p_n\}$, such that $\eta_{\widetilde p_n} \stackrel{a.s.}{\to} \eta$ almost surely, as $\widetilde p_n\to
\infty$.
\end{lemma}

\item Prove that the Benjamini-Hochberg procedure in Definition \ref{def:BH} controls FDR for independent data.\\

{\em Hint:} \fbox{ Give a good hint. }

\item  
\end{enumerate}

### 2020-08-20 abstract.tex
minor updates
going back to the original title
addded some exercises
to Chapter 2.

This text presents a collection of new results and recent developments on the phase transition phenomena in sparse signal problems.
The main theme is the study of the fundamental limits in high-dimensional testing and inference. Since the seminal works
of \cite{ingster1998minimax} and \cite{donoho2004higher}, the subject has received a lot of attention in the literature with important contributions
from \fbox{Add references}.  These works (among others) have discovered the presence of fundamental limits in the so-called 
{\em needle in haystack} problems, where a sparse signal is observed in high-dimensional additive noise.  In this setting, two archetypal 
problems arise -- the {\em signal detection} and {\em support support recovery}.  Signal detection is a hypothesis testing problem, which amounts
to testing for the presence of non-zero signal in some (any) of the dimensions.  Support recovery, on the other hand, is an inference problem, which aims to estimate the signal support -- the locations of the non-zero signal components.  The fundamental limits of these problems are studied in the so-called high-dimensional asymptotics regime where the sample size $n$ is fixed and the dimension $p$ of the underlying signal 
grows to infinity.

From a probabilistic perspective, the above phase transition phenomena are asymptotic zero-one type laws, as $p\to\infty$.  
Namely, consider a sparse signal with  {\em support size} on the order of $p^{1-\beta}$ for some parameter $\beta\in (0,1)$.  
Parameterize the non-zero {\em signal amplitude} by $\Delta(p^r)$, 
for some $r>0$ and a suitable monotone non-decreasing function $\Delta(\cdot)$.  Then, for a broad range of error distributions and
statistical problems, one encounters a sharp transition between the regimes where the problem is solvable and unsolvable depending on 
the signal magnitude $r$ and signal sparsity $\beta$.  More precisely, there exists a boundary function $f(\beta)$ such that if 
the signal magnitudes are {\em above} the boundary, $r> f(\beta)$, then the problem can be solved with vanishing loss as $p\to\infty$, with a suitable statistical procedure.  On the other hand, if the signal is below that same boundary, i.e., $r<f(\beta)$, all statistical procedures fail to  provide a solution with a vanishing loss, as $p\to\infty$. Of course, depending on whether one considers the  detection (testing) or support recovery 
(inference) problems, different loss functions quantify success and failure. The choice of the loss functions is often guided by the 
applications, resulting in a wide variety of phase-transitions (see e.g.\ Figure \ref{FIGURE} \fbox{FIGURE?}).

{\bf The contributions of this work.} The fundamental limits of the classic detection problem hinge 
of the analysis of the discrepancy between the {\em null} and {\em alternative} hypotheses e.g., via Hellinger distance.  Thus, perhaps 
for  technical reasons, much of the analysis was done under the assumption that the additive errors are independent and/or Gaussian.  
In this work, we demonstrate that the support recovery problems, especially {\em exact support recovery}, are best understood from 
the novel perspective of the {\em concentration of maxima} phenomenon in extreme value theory.  It turns out that under a very broad 
range of light-tailed error distributions and under a {\em very} broad range of error dependence structures, the maxima of the errors, 
when rescaled (but not centered!) converge in probability to a positive constant. This concentration property leads to a complete solution 
of the exact  support recovery problem for the broad family of thresholding procedures.  Most if not all existing support estimation procedures 
are types thresholding procedures.   That is, the signal support estimate comprises of all components exceeding a suitable (potentially 
data-dependent)  threshold.  We show, by exploiting concentration of maxima, that thresholding procedures obey a phase-transition, where if the signal is above a 
certain boundary, asymptotically exact recovery is possible while below the boundary all thresholding procedures fail, as $p\to\infty$.  
Remarkably, light-tailed maxima concentrate under very broad and strong dependence.  This is exemplified by our characterization of
the concentration of maxima phenomenon for Gaussian triangular arrays.  For example, in the special case of stationary Gaussian time
series, vanishing auto-covariance is necessary and sufficient for the maxima to concentrate in the same way as independent standard 
normal random variables.   These probabilistic contributions are of independent interest and extend classic work of \cite{berman1964limit}. 
Concentration of maxima is a type of super-concentration phenomenon studied also in \cite{chatterjee:2014} and \cite{tanguy:2015}.
The robustness of the concentration of maxima phenomenon to dependence can perhaps explain the universality 
of phase transitions in support recovery problems.

The use of concentration of maxima phenomenon highlights one core idea in our work, which allows for a first of its kind comprehensive 
treatment of thresholding procedures under very broad error-dependence conditions.  The text involves also a full spectrum of related results 
such as minimax-optimality and finite-sample Bayes-optimality in support estimation.  Using different type of loss functions and typeI error 
controls, we obtain a rich picture of the exact and approximate support recovery problems in high dimensions.  
Many of these phase transitions have not appeared in previously published literature.

High-dimensional support recovery problems arise in many modern applications ranging from cybersecurity to  statistical genetics.  
Genome-wide association studies (GWAS) in genetics is a particularly natural application, where the theoretical phase-transition results help 
explain and quantify a previously observed empirical phenomenon of the so-called steep part of the power curve.  
In the last chapter of this work, we detail this application and highlight future theoretical and practical consequences of our work.

@@

{\bf Acknowledgements.} The authors gratefully acknowledge the support of all colleagues from the Statistics Department at the 
University of Michigan, Ann Arbor.   Special thanks (in alphabetical order) go to Xuming He, Tailen Hsing, Michalis Kallitsis, Liza Levina, 
Rodderick Little, Ya'acov Ritov, Kerby Shedden, Jinqi Shen, Jonathan Terhorst, Gongjun Xu, \fbox{add more}
The authors were partially supported by the NSF program {\em Algorithms for Threat Detection}.

>>>>

This text presents a collection of new results and recent developments on the phase transition phenomena in sparse signal problems.
The main theme is the study of the fundamental limits in high-dimensional testing and inference. Since the seminal works
of \cite{ingster1998minimax} and \cite{donoho2004higher}, the subject has received a lot of attention in the literature with important contributions
from \cite{ji2012ups,genovese2012comparison,jin2014optimality,arias2017distribution,butucea2018variable}.  These works (among others) 
 have discovered the presence of fundamental limits in the so-called  {\em needle in haystack} problems, where a sparse signal is observed in 
 high-dimensional additive noise.  In this setting, two archetypal  problems arise -- the {\em signal detection} and {\em support support recovery}.  
 Signal detection is a hypothesis testing problem, which amounts
to testing for the presence of non-zero signal in some (any) of the dimensions.  Support recovery, on the other hand, is an inference problem, which aims to estimate the signal support -- the locations of the non-zero signal components.  The fundamental limits of these problems are studied in the so-called high-dimensional asymptotics regime where the sample size $n$ is fixed and the dimension $p$ of the underlying signal 
grows to infinity.

From a probabilistic perspective, the above phase transition phenomena are asymptotic zero-one type laws, as $p\to\infty$.  
Namely, consider a sparse signal with  {\em support size} on the order of $p^{1-\beta}$ for some parameter $\beta\in (0,1)$.  
Parameterize the non-zero {\em signal amplitude} by $\Delta(p^r)$, 
for some $r>0$ and a suitable monotone non-decreasing function $\Delta(\cdot)$.  Then, for a broad range of error distributions and
statistical problems, one encounters a sharp transition between the regimes where the problem is solvable and unsolvable depending on 
the signal magnitude $r$ and signal sparsity $\beta$.  More precisely, there exists a boundary function $f(\beta)$ such that if 
the signal magnitudes are {\em above} the boundary, $r> f(\beta)$, then the problem can be solved with vanishing loss as $p\to\infty$, with a suitable statistical procedure.  On the other hand, if the signal is below that same boundary, i.e., $r<f(\beta)$, all statistical procedures fail to  provide a solution with a vanishing loss, as $p\to\infty$. Of course, depending on whether one considers the  detection (testing) or 
support recovery (inference) problems, different loss functions quantify success and failure. The choice of the loss functions is often guided by the 
applications, resulting in a rich picture of phase-transitions (see e.g.\ Figure \ref{fig:phase-Gaussian-errors}).

{\bf The contributions of this work.} The fundamental limits of the classic detection problem hinge 
of the analysis of the discrepancy between the {\em null} and {\em alternative} hypotheses e.g., via Hellinger distance.  Thus, perhaps 
for  technical reasons, much of the analysis in the existing literature has been done under the assumption that the additive errors are 
independent and/or Gaussian, or using loss functions unaffected by the dependence such as the Hamming loss.  
In this work, we demonstrate that the support recovery problems, especially {\em exact support recovery}, are best understood from 
the novel perspective of the {\em concentration of maxima} phenomenon in extreme value theory.  It turns out that under a very broad 
range of light-tailed error distributions and under a {\em very} broad range of error dependence structures, the maxima of the errors, 
when rescaled (but not centered!) converge in probability to a positive constant. This concentration property leads to a complete solution 
of the exact  support recovery problem for the broad family of thresholding procedures.  Most if not all existing support estimation 
procedures are types thresholding procedures (see Section \ref{sec:statistical-procedures}).   
That is, the signal support estimate comprises of all components exceeding a suitable (potentially data-dependent)  threshold.  We show, by exploiting concentration of maxima, that thresholding procedures obey a phase-transition, where if the signal is above a 
certain boundary, asymptotically exact recovery is possible while below the boundary all thresholding procedures fail, as $p\to\infty$.  
Remarkably, light-tailed maxima concentrate under very broad and strong dependence.  This is exemplified by our characterization of
the concentration of maxima phenomenon for Gaussian triangular arrays.  For example, in the special case of stationary Gaussian time
series, vanishing auto-covariance is necessary and sufficient for the maxima to concentrate in the same way as independent standard 
normal random variables.   These probabilistic contributions are of independent interest and extend classic work of \cite{berman1964limit}. 
Concentration of maxima is a type of super-concentration phenomenon studied also in \cite{chatterjee2014superconcentration} 
and \cite{tanguy:2015}. The robustness of the concentration of maxima phenomenon to dependence can perhaps explain the universality 
of phase transitions in support recovery problems.

The use of concentration of maxima phenomenon highlights one core idea in our work, which allows for a first of its kind comprehensive 
treatment of thresholding procedures under very broad error-dependence conditions.  The text involves also a full spectrum of related results 
such as minimax-optimality and finite-sample Bayes optimality in support estimation.  Using different type of loss functions and type I error 
controls, we obtain a rich picture of the exact and approximate support recovery problems in high dimensions.  
Many of these phase transition results have not appeared in previously published literature.

High-dimensional support recovery problems arise in many modern applications ranging from cybersecurity, theoretical computer science,
to statistical genetics.  Genome-wide association studies (GWAS) in genetics is a particularly natural application, where the asymptotic 
phase-transition results help explain and quantify a previously observed empirical phenomenon of the so-called steep part of the power 
curve.  In the last chapter of this work, we detail this application and highlight future theoretical and practical consequences of our work.

@@

{\bf Acknowledgements.} The authors gratefully acknowledge the support of their families and all colleagues from the Statistics Department 
at the University of Michigan, Ann Arbor.   Special thanks (in alphabetical order) go to Xuming He, Tailen Hsing, Michalis Kallitsis, Liza Levina, 
Rodderick Little, Ya'acov Ritov, Kerby Shedden, Jinqi Shen, Jonathan Terhorst, Gongjun Xu, \fbox{add more}
The authors were partially supported by the NSF program {\em Algorithms for Threat Detection}.
